---
title: "Assignment 4 - HPC, SQL, and interactive plots"
author: "Nate Truettner"
format: html
---
```{r}
library(microbenchmark)
library(ggplot2)
library(plotly)
library(dplyr)
```

## Question 1
```{r}
# running cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

fun2alt <- function(mat) {
  ans <- mat
  for (j in 2:ncol(mat)) {
    ans[, j] <- ans[, j] + ans[, j - 1]
  }
  ans
}
```

```{r}
# Use the functions with this data
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)
identical(fun2(dat), fun2alt(dat)) 
mb <- microbenchmark(
  original   = fun2(dat),
  updated = fun2alt(dat),
  times = 100L
)
mb
autoplot(mb) +
  labs(title = "fun2 vs fun2alt runtime", x = "Time (µs)") +
  theme_minimal()
```
As shown in the plot, we notice that the updated function takes significantly less time to run than the original function. By cutting down to only one for loop, we greatly enhance the efficiency of the process.

## Question 2
```{r}
fun3 <- function(x) {
  out <- numeric(length(x))
  cumsum_x <- 0
  for (i in seq_along(x)) {
    if (x[i] == 0) {
      cumsum_x <- 0
    } else {
      cumsum_x <- cumsum_x + x[i] 
    }
    out[i] <- cumsum_x
  }
  out
}

library(Rcpp)

cppFunction('
NumericVector fast_fun3_cpp(NumericVector x) {
  int n = x.size();
  NumericVector out(n);
  
  double cs = 0.0;
  
  for (int i = 0; i < n; i++) {
    if (x[i] == 0) {
      cs = 0.0;
    } else {
      cs += x[i];
    }
    out[i] = cs;
  }
  
  return out;
}
')
set.seed(123)
x <- sample(c(0:3), 1e6, replace = TRUE)


mb <- microbenchmark(
  original = fun3(x),
  cpp      = fast_fun3_cpp(x),
  times = 50
)
print(mb)
```
I tried many different implementations only using R, so I looked up how to use Rcpp to speed up the process and found that it significantly sped up the function.

## Question 3
```{r}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132

library(parallel)

set.seed(1231)

system.time({
  cl <- makeCluster(detectCores() - 1)
  
  clusterExport(cl, varlist = c("sim_pi"))
  
  ans_par <- unlist(parLapply(cl, 1:4000, sim_pi, n = 10000))
  
  stopCluster(cl)
  
  print(mean(ans_par))
})

#now run for initial process
set.seed(1231)

system.time({
  ans_seq <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans_seq))
})

library(ggplot2)

df <- data.frame(
  value = c(ans_seq, ans_par),
  method = rep(c("Sequential", "Parallel"), each = length(ans_seq))
)

ggplot(df, aes(x = value, fill = method)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = pi, linetype = "dashed") +
  labs(
    title = "Distribution of Pi Estimates",
    x = "Estimated Pi",
    y = "Density"
  ) +
  theme_bw()


mb <- microbenchmark(
  sequential = unlist(lapply(1:4000, sim_pi, n = 10000)),
  parallel = {
    cl <- makeCluster(detectCores() - 1)
    clusterExport(cl, "sim_pi")
    res <- unlist(parLapply(cl, 1:4000, sim_pi, n = 10000))
    stopCluster(cl)
    res
  },
  times = 10
)

print(mb)

autoplot(mb)
```
While it is not a dramatic improvement, the parallel process did slightly speed up the computation. I wonder if my computer being old and not very powerful made this closer than I thought it would be. Both methods yield similar results for estimation of pi.

## Question 4
```{r}
library(RSQLite)
library(DBI)

con <- dbConnect(RSQLite::SQLite(), dbname = ":memory:")

surveys <- read.csv("surveys.csv")
species <- read.csv("species.csv")
plots   <- read.csv("plots.csv")

dbWriteTable(con, "surveys", surveys)
dbWriteTable(con, "species", species)
dbWriteTable(con, "plots",   plots)

dbListTables(con)
```
```{sql connection=con}
PRAGMA table_info(surveys)
```
```{sql connection=con}
SELECT month, day, year, species_id, ROUND(weight / 1000.0, 2) AS weight_kg
FROM surveys
WHERE year = 1999
ORDER BY species_id
```
## Question 5
```{sql connection=con}
SELECT SUM(weight) AS "Total Weight", AVG(weight) AS "Average Weight", MIN(weight) AS "Minimum Weight", MAX(weight) AS "Maximum Weight"
FROM surveys
WHERE weight BETWEEN 5 AND 10
```
## Question 6
```{r}
yearly_species_summary <- dbGetQuery(con, "
  SELECT year,
         species_id,
         COUNT(*)       AS count,
         AVG(weight)    AS avg_weight
  FROM surveys
  WHERE species_id IS NOT NULL
  GROUP BY year, species_id
  ORDER BY year, species_id
")

yearly_species_summary
```
## Question 7
```{sql connection=con}
SELECT taxa, COUNT(*) AS "Number of Species"
FROM species
GROUP BY taxa
HAVING COUNT(*) > 10
```
## Question 8
```{r}
genus_plot_counts <- dbGetQuery(con, "
  SELECT p.plot_id,
         sp.genus,
         COUNT(*) AS num_animals
  FROM surveys s
  JOIN plots   p  ON s.plot_id   = p.plot_id
  JOIN species sp ON s.species_id = sp.species_id
  WHERE sp.genus IS NOT NULL
  GROUP BY p.plot_id, sp.genus
  ORDER BY p.plot_id ASC, num_animals DESC
")

genus_plot_counts
```
## Question 9
```{sql connection=con}
SELECT p.plot_type,
         sp.species_id,
         sp.genus,
         sp.species,
         COUNT(*) AS "Number Captured"
FROM surveys s
JOIN plots   p  ON s.plot_id    = p.plot_id
JOIN species sp ON s.species_id = sp.species_id
WHERE sp.species != 'sp.' AND sp.species IS NOT NULL
GROUP BY p.plot_type, sp.species_id, sp.genus, sp.species
ORDER BY p.plot_type, "Number Captured" DESC
```
## Question 10
```{sql connection=con}
SELECT sp.species_id,
         sp.species,
         AVG(s.weight) AS avg_weight
FROM surveys s
JOIN species sp ON s.species_id = sp.species_id
WHERE sp.taxa = 'Rodent' AND s.weight IS NOT NULL
GROUP BY sp.species_id
ORDER BY avg_weight DESC
```
## Question 11
```{r}
library(ggplot2)
library(plotly)

plot_subset <- subset(genus_plot_counts, plot_id %in% 1:5)
plot_subset$plot_id <- as.factor(plot_subset$plot_id)

p <- ggplot(plot_subset, aes(x = plot_id, y = num_animals, fill = genus)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Number of Animals Captured by Genus and Plot (Plots 1-5)",
    x     = "Plot ID",
    y     = "Number of Animals",
    fill  = "Genus"
  ) +
  theme_minimal()

ggplotly(p)
```
## Question 12
```{r}
library(dplyr)

smoothed <- yearly_species_summary %>%
  filter(!is.na(avg_weight)) %>%
  group_by(species_id) %>%
  arrange(year) %>%
  mutate(smooth_weight = zoo::rollmean(avg_weight, k = 3, fill = NA, align = "center")) %>%
  filter(!is.na(smooth_weight)) %>%
  ungroup()

plot_ly(
  data   = smoothed,
  x      = ~year,
  y      = ~smooth_weight,
  color  = ~species_id,
  type   = "scatter",
  mode   = "lines",
  text   = ~paste("Species:", species_id,
                  "<br>Year:", year,
                  "<br>Avg Weight (smoothed):", round(smooth_weight, 2)),
  hoverinfo = "text"
) %>%
  layout(
    title = list(text = "Average Weight Over Time by Species (3-Year Rolling Mean)"),
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average Weight (g)"),
    legend = list(title = list(text = "Species")),
    hovermode = "closest"
  )
```








