---
title: "Homework 1 - Exploratory Data Analysis"
format: html
author: "Nate Truettner"
embed-resources: TRUE
---

## 1
```{r}
library(dplyr)
library(lubridate)
pm_2005 <- read.csv('pm_25_mich_2005.csv')
pm_2025 <- read.csv('pm_25_mich_2025.csv')
```

### 2005 data
```{r}
dim(pm_2005)
head(pm_2005)
tail(pm_2005)
str(pm_2005)
hist(pm_2005$Daily.Mean.PM2.5.Concentration)
```
### 2025 data
```{r}
dim(pm_2025)
head(pm_2025)
tail(pm_2025)
str(pm_2025)
hist(pm_2025$Daily.Mean.PM2.5.Concentration, xlim=c(0,80))
```

The 2005 dataset has 5280 rows and 22 columns while the 2025 dataset has 9934 rows and 22 columns. The datasets appear to share the same columns, which include integer, numeric, and character type variables. Assessing the distribution of PM2.5 in 2005 and 2025, we notice that the 2005 dataset has a more prominent tail, meaning that there were more recordings of high PM2.5 in 2005 than 2025. Additionally, the mean PM2.5 value seems to be lower in 2025 as the first bin (assume low PM2.5) is a very tall peak. From this, we can reasonably conclude that PM2.5 has generally decreased since 2005 in the state of Michigan.

## 2
```{r}
full_pm <- rbind(pm_2005, pm_2025)
full_pm <- full_pm %>%
  rename(pm2.5 = Daily.Mean.PM2.5.Concentration,
         AQI = Daily.AQI.Value)
full_pm <- full_pm %>%
  mutate(Date = mdy(Date))
full_pm <- full_pm %>%
  mutate(year = year(Date))
str(full_pm)
```
## 3
```{r}
library(leaflet)
library(ggplot2)
library(maps)
us_map <- map_data("state")
mi_map <- map_data("state") %>%
  filter(region == "michigan")
ggplot() +
  geom_polygon(data = mi_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "white") +
  geom_point(data = full_pm,
             aes(x = Site.Longitude, y = Site.Latitude),
             color = "steelblue",
             alpha = 0.7) +
  coord_fixed(1.3) +
  facet_wrap(~year) +
  labs(title = "Michigan PM2.5 Monitoring Sites: 2005 vs 2025") +
  theme_minimal()

```
The 2005 distribution shows that the monitoring sites are mostly concentrated in the southern half of Michigan, with many in the southeast. In 2025, we notice that a few of the sites in the upper peninsula are no longer active, but there are a comparable number in the mainland. Generally, there appear to be fewer monitoring sites in 2025 compared to 2005.

## 4
```{r}
summary(full_pm$pm2.5)
```
Thankfully, there are no NA values for the PM2.5 variable in the combined dataset. The values range from 0 to 116, with a median of 7.4 ug/m3 and a mean of 10.4. The maximum value of 116 might be an outlier. I did some research and apparently you can get a value of more than 100 if there's an active fire; it's possible there was a forest fire near the site that could have influenced this observation. I don't think there are any values that are implausible.

```{r}
table(full_pm$Method.Code)
```
The two most common method codes are 118 and 636. Code 118 corresponds to gravimetric analysis, which is done intermittently while code 636 corresponds to a continuous air monitor performing broadband spectroscopy.

```{r}
table(pm_2005$Method.Code)
table(pm_2025$Method.Code)
```
Looking at data from each year, it is evident that in 2005, code 118 was prevalent while in 2025 636 is most common. This represents a shift from manual lab analysis to automated data collection. This is notable as differences over time may not be true differences in PM2.5, but simply reflecting the adjustment from one collection technique to another. There are no missing values for PM2.5 in the dataset.

## 5

### State-level
```{r}
state_summary <- full_pm %>%
  group_by(year) %>%
  summarise(
    mean_pm25 = mean(pm2.5),
    median_pm25 = median(pm2.5),
    max_pm25 = max(pm2.5),
    min_pm25 = min(pm2.5)
  )
state_summary

ggplot(state_summary, aes(x = factor(year), y = mean_pm25, fill = "Mean")) +
  geom_bar(stat = "identity") +
  labs(
    title = "State-Level PM2.5 Trends in Michigan (2005–2025)",
    x = "Year",
    y = "Mean PM2.5 (µg/m³)",
    fill = "Statistic"
  ) +
  theme_minimal()
```
At the state level, PM2.5 appears to have decreased in the past 20 years. The mean PM2.5 value in 2005 was 13.77 ug/m3 while in 2025 it was 8.30. The range of observations is greater in 2025 than 2005, suggesting possible larger variance in the automated collection system.

### County-level
```{r}
county_summary <- full_pm %>%
  group_by(County, year) %>%
  summarise(mean_pm25 = mean(pm2.5),
            median_pm25 = median(pm2.5),
            latitude = mean(Site.Latitude),
            longitude = mean(Site.Longitude))

county_summary

ggplot(county_summary, aes(x = year, y = mean_pm25, color = County)) +
  geom_line(alpha = 0.7) +
  labs(
    title = "County-Level PM2.5 Trends in Michigan",
    x = "Year",
    y = "Mean PM2.5 (µg/m³)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplot() +
  geom_polygon(data = mi_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "white") +
  geom_point(data = county_summary,
             aes(x = longitude, y = latitude, color = mean_pm25),
             size = 4, alpha = 0.8) +
  scale_color_viridis_c(option = "C") +
  facet_wrap(~year) +
  coord_fixed(1.3) +
  labs(title = "County-Level Mean PM2.5 Concentration in Michigan",
       color = "Mean PM2.5 (ug/m3)") +
  theme_minimal()
```
At the county level, I first made a line/spaghetti plot to see if mean PM2.5 has generally decreased in every county. In almost every county, PM2.5 concentration decreased in 2025 from 2005. The only exception is Schoolcraft County, which saw about a 1 ug/m3 increase in PM2.5. Examining the spatial plot with a color gradient describing PM2.5 concentration, we notice that generally, PM2.5 has gone down over time, as the hue turns from yellow to purple.


### City-level
```{r}
wayne <- full_pm %>%
  filter(County == "Wayne")
table(wayne$Local.Site.Name)
wayne_summary <- wayne %>%
  group_by(Local.Site.Name, year) %>%
  summarise(
    mean_pm25 = mean(pm2.5),
    median_pm25 = median(pm2.5),
    latitude = mean(Site.Latitude),
    longitude = mean(Site.Longitude)
  )

ggplot(wayne_summary, aes(x = year, y = mean_pm25, color = Local.Site.Name)) +
  geom_line(alpha = 0.7) +
  labs(
    title = "Wayne County PM2.5 Trends by Site",
    x = "Year",
    y = "Mean PM2.5 (µg/m³)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplot() +
  geom_polygon(data = mi_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "white") +
  geom_point(data = wayne_summary,
             aes(x = longitude, y = latitude, color = mean_pm25),
             size = 4, alpha = 0.8) +
  coord_fixed(1.3, xlim = c(-83.5, -82.7), ylim = c(42.0, 42.5)) +
  scale_color_viridis_c(option = "C") +
  facet_wrap(~year) +
  labs(title = "Wayne County Mean PM2.5 Concentration",
       x = "Longitude",
       y = "Latitude",
       color = "Mean PM2.5") +
  theme_minimal()
```
I wanted to perform a similar analysis within Wayne County that I performed for the county-level analysis. Looking at the line plot, we see that every site reported decreased PM2.5 concentration in 2025, and the lines are essentially parallel. Drops in PM2.5 do not appear to be constrained to specific sites. The spatial plot is complicated a bit by some sites only active in 2005 or 2025, but we see generally that mean PM2.5 concentration has decreased throughout the county as the hue has turned into a deep purple at most sites.











