---
title: "Homework 2"
format: html
author: "Nate Truettner"
embed-resources: TRUE
---

## Load in data
```{r}
library(tidyverse)
chs_individual <- read_csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/chs/chs_individual.csv")
chs_regional <- read_csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/chs/chs_regional.csv")
```
## 1
```{r}
dim(chs_individual)
dim(chs_regional)
chs_merged <- chs_individual %>%
  left_join(chs_regional, by = "townname")
dim(chs_merged)
```
The number of rows in chs_merged and chs_individual match, so we can proceed.

```{r}
#Handle missing data
library(DescTools)
str(chs_merged)
continuous_vars <- c("agepft", "height", "weight", "bmi", "fev", "fvc", "mmef",
                     "pm25_mass", "pm25_so4", "pm25_no3", "pm25_nh4", "pm25_oc",
                     "pm25_ec", "pm25_om", "pm10_oc", "pm10_ec", "pm10_tc",
                     "formic", "acetic", "hcl", "hno3", "o3_max", "o3106",
                     "o3_24", "no2", "pm10", "no_24hr", "pm2_5_fr",
                     "iacid", "oacid", "total_acids")
categorical_vars <- c("asthma", "active_asthma", "father_asthma", "mother_asthma",
                      "wheeze", "hayfever", "allergy", "educ_parent", 
                      "smoke", "pets", "gasstove", "townname", "race")

chs_merged <- chs_merged %>%
  dplyr::group_by(male, hispanic) %>%
  mutate(
    across(all_of(continuous_vars), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)),
    across(all_of(categorical_vars), ~ ifelse(is.na(.), Mode(., na.rm = TRUE)[1], .))
  ) %>%
  ungroup()

sum(is.na(chs_merged))
head(chs_merged)
```
## 2
```{r}
chs_merged <- chs_merged %>%
  mutate(obesity_level = case_when(bmi < 14 ~ "underweight",
                                   bmi >= 14 & bmi < 22 ~ "normal",
                                   bmi >= 22 & bmi < 24 ~ "overweight",
                                   bmi >= 24 ~ "obese" ))
chs_merged %>%
  dplyr::group_by(obesity_level) %>%
  dplyr::summarise(
    min_bmi = min(bmi, na.rm = TRUE),
    max_bmi = max(bmi, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(min_bmi)
```
## 3
```{r}
chs_merged <- chs_merged %>%
  mutate(smoke_gas_exposure = case_when(
    smoke == 0 & gasstove == 0 ~ "No exposure",
    smoke == 1 & gasstove == 0 ~ "Smoke only",
    smoke == 0 & gasstove == 1 ~ "Gas stove only",
    smoke == 1 & gasstove == 1 ~ "Both"
  ))

table(chs_merged$smoke_gas_exposure)
```
## 4
```{r}
fev1_town <- chs_merged %>%
  dplyr::group_by(townname) %>%
  dplyr::summarise(
    mean_fev = mean(fev, na.rm = TRUE),
    sd_fev   = sd(fev, na.rm = TRUE),
    n        = n()
  ) %>%
  arrange(desc(mean_fev))
fev1_town

fev1_sex <- chs_merged %>%
  dplyr::group_by(male) %>%
  dplyr::summarise(
    mean_fev = mean(fev, na.rm = TRUE),
    sd_fev   = sd(fev, na.rm = TRUE),
    n        = n()
  ) %>%
  mutate(male = ifelse(male == 1, "Male", "Female"))
fev1_sex

fev1_obesity <- chs_merged %>%
  dplyr::group_by(obesity_level) %>%
  dplyr::summarise(
    mean_fev = mean(fev, na.rm = TRUE),
    sd_fev   = sd(fev, na.rm = TRUE),
    n        = n()
  ) %>%
  arrange(desc(mean_fev))
fev1_obesity

fev1_smoke_gas <- chs_merged %>%
  dplyr::group_by(smoke_gas_exposure) %>%
  dplyr::summarise(
    mean_fev = mean(fev, na.rm = TRUE),
    sd_fev   = sd(fev, na.rm = TRUE),
    n        = n()
  ) %>%
  arrange(desc(mean_fev))
fev1_smoke_gas
```
Looking at the distinct cities, Alpine has the highest mean FEV1 value while Mira Loma has the lowest. On average, males have a higher FEV1 value than females, which is perhaps worth exploring. FEV1 also seems to scale with BMI, as a higher BMI is associated with a higher FEV1. Lastly, and perhaps interestingly, having neither smoke nor gas stove exposure is associated with the highest average FEV1 value.

## Looking at the data (EDA)
```{r}
library(patchwork)
# Checking variable types
dim(chs_merged)

chs_merged %>%
  dplyr::select(bmi, fev, smoke_gas_exposure, pm25_mass, obesity_level) %>%
  str()

# Check for NAs
chs_merged %>%
  dplyr::select(bmi, fev, smoke_gas_exposure, pm25_mass, obesity_level) %>%
  is.na() %>%
  colSums()

#no NAs remaining

# Summary stats
chs_merged %>%
  dplyr::select(bmi, fev, pm25_mass) %>%
  summary()

table(chs_merged$obesity_level)
table(chs_merged$smoke_gas_exposure)

#nothing looks out of place

# Distributions of continuous variables
p1 <- ggplot(chs_merged, aes(x = bmi)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(title = "Distribution of BMI", x = "BMI", y = "Count") +
  theme_minimal()

p2 <- ggplot(chs_merged, aes(x = fev)) +
  geom_histogram(bins = 30, fill = "firebrick", alpha = 0.8) +
  labs(title = "Distribution of FEV", x = "FEV (ml)", y = "Count") +
  theme_minimal()

p3 <- ggplot(chs_merged, aes(x = pm25_mass)) +
  geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.8) +
  labs(title = "Distribution of PM2.5", x = "PM2.5 (µg/m³)", y = "Count") +
  theme_minimal()

vars <- p1 | p2 | p3
vars

# BMI vs. FEV

# Scatterplot with regression line
ggplot(chs_merged, aes(x = bmi, y = fev)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE) +
  labs(
    title = "Association Between BMI and FEV",
    x = "BMI",
    y = "FEV (ml)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.line = element_line(color = "black"))

# Correlation
cor(chs_merged$bmi, chs_merged$fev, use = "complete.obs")

# Boxplot by obesity category
ggplot(chs_merged, aes(x = obesity_level, y = fev, fill = obesity_level)) +
  geom_boxplot(alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "white", color = "black") +
  scale_fill_brewer(palette = "RdYlGn") +
  scale_x_discrete(limits = c("underweight", "normal", "overweight", "obese")) +
  labs(
    title = "FEV by Obesity Level",
    x = "Obesity Level",
    y = "FEV (ml)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        axis.line = element_line(color = "black"))

# Smoke/gas exposure vs. FEV

ggplot(chs_merged, aes(x = smoke_gas_exposure, y = fev, fill = smoke_gas_exposure)) +
  geom_boxplot(alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "white", color = "black") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "FEV by Smoke and Gas Exposure",
    x = "Exposure Type",
    y = "FEV (ml)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        axis.line = element_line(color = "black"),
        axis.text.x = element_text(angle = 15, hjust = 1))

# Summary table
chs_merged %>%
  dplyr::group_by(smoke_gas_exposure) %>%
  dplyr::summarise(
    mean_fev = round(mean(fev, na.rm = TRUE), 1),
    sd_fev   = round(sd(fev, na.rm = TRUE), 1),
    n        = n()
  )

# PM2.5 vs FEV

# Individual level
ggplot(chs_merged, aes(x = pm25_mass, y = fev)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE) +
  labs(
    title = "PM2.5 vs FEV (Individual Level)",
    x = "PM2.5 (µg/m³)",
    y = "FEV (ml)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.line = element_line(color = "black"))

# Town-level aggregation
chs_merged %>%
  dplyr::group_by(townname) %>%
  dplyr::summarise(
    mean_pm25 = mean(pm25_mass, na.rm = TRUE),
    mean_fev  = mean(fev, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = mean_pm25, y = mean_fev, label = townname)) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(vjust = -0.7, size = 3) +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE) +
  labs(
    title = "Town-Level Mean PM2.5 vs Mean FEV",
    x = "Mean PM2.5 (µg/m³)",
    y = "Mean FEV (ml)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.line = element_line(color = "black"))

# Correlation at town level
chs_merged %>%
  dplyr::group_by(townname) %>%
  dplyr::summarise(
    mean_pm25 = mean(pm25_mass, na.rm = TRUE),
    mean_fev  = mean(fev, na.rm = TRUE)
  ) %>%
  summarise(cor = cor(mean_pm25, mean_fev))
```

An increase in BMI appears to be associated with an increase in FEV. From the boxplot, it does not appear that FEV varies greatly across levels of smoke/gas exposure. Assessing the scatterplot of FEV vs PM2.5 at the town-level, it seems like a higher PM2.5 exposure is associated with lower FEV.


## Visualizations
```{r}
library(ggplot2)
library(leaflet)

#Facet plot
ggplot(chs_merged, aes(x = bmi, y = fev)) +
  geom_point(alpha = 0.3, size = 0.8, color = "steelblue") +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE) +
  facet_wrap(~ townname, ncol = 4) +
  labs(
    title = "BMI vs FEV by Town",
    x = "BMI",
    y = "FEV (mL)"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.line = element_line(color = "black"))

#Stacked histogram: FEV by BMI category
ggplot(chs_merged, aes(x = fev, fill = obesity_level)) +
  geom_histogram(bins = 30, position = "stack", alpha = 0.8) +
  scale_fill_brewer(palette = "RdYlGn") +
  labs(
    title = "Distribution of FEV by BMI Category",
    x = "FEV (ml)",
    y = "Count",
    fill = "BMI Category"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.line = element_line(color = "black"))

#Stacked histogram: FEV by smoke/gas exposure
ggplot(chs_merged, aes(x = fev, fill = smoke_gas_exposure)) +
  geom_histogram(bins = 30, position = "stack", alpha = 0.8) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Distribution of FEV by Smoke/Gas Exposure",
    x = "FEV (ml)",
    y = "Count",
    fill = "Exposure"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.line = element_line(color = "black"))

#Bar chart of BMI by smoke/gas exposure
ggplot(chs_merged, aes(x = smoke_gas_exposure, fill = obesity_level)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "RdYlGn") +
  labs(
    title = "BMI Category by Smoke/Gas Exposure",
    x = "Smoke/Gas Exposure",
    y = "Count",
    fill = "BMI Category"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.line = element_line(color = "black"),
    axis.text.x = element_text(angle = 15, hjust = 1)
  )

#Statistical summary graphs of FEV by BMI and FEV by smoke/gas exposure category
ggplot(chs_merged, aes(x = obesity_level, y = fev, fill = obesity_level)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "white", color = "black") +
  scale_fill_brewer(palette = "RdYlGn") +
  labs(
    title = "FEV by BMI Category",
    x = "BMI Category",
    y = "FEV (ml)",
    fill = ""
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        axis.line = element_line(color = "black"))

ggplot(chs_merged, aes(x = smoke_gas_exposure, y = fev, fill = smoke_gas_exposure)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "white", color = "black") +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "FEV by Smoke/Gas Exposure",
    x = "Exposure",
    y = "FEV (ml)",
    fill = ""
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        axis.line = element_line(color = "black"),
        axis.text.x = element_text(angle = 15, hjust = 1))

#A leaflet map showing the concentrations of PM2.5 mass in each of the CHS communities
pm25_towns <- chs_merged %>%
  dplyr::group_by(townname) %>%
  dplyr::summarise(
    pm25  = mean(pm25_mass, na.rm = TRUE),
    lat   = mean(lat, na.rm = TRUE),
    lon   = mean(lon, na.rm = TRUE)
  )

pal <- colorNumeric(palette = "YlOrRd", domain = pm25_towns$pm25)

leaflet(pm25_towns) %>%
  addTiles() %>%
  addCircleMarkers(
    lng    = ~lon,
    lat    = ~lat,
    radius = ~pm25 / 2,
    color  = ~pal(pm25),
    fillOpacity = 0.8,
    label  = ~paste0(townname, ": ", round(pm25, 1), " µg/m³")
  ) %>%
  addLegend(
    position = "bottomright",
    pal      = pal,
    values   = ~pm25,
    title    = "PM2.5 (µg/m³)"
  )

# Is PM2.5 associated with FEV?
chs_merged %>%
  dplyr::group_by(townname) %>%
  dplyr::summarise(
    mean_pm25 = mean(pm25_mass, na.rm = TRUE),
    mean_fev  = mean(fev, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = mean_pm25, y = mean_fev, label = townname)) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(vjust = -0.7, size = 3) +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE) +
  labs(
    title = "Town-Level PM2.5 vs Mean FEV",
    x     = "Mean PM2.5 (µg/m³)",
    y     = "Mean FEV (ml)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.line = element_line(color = "black"))
```
From our final visualization, it appears that a higher PM2.5 value is associated with a lower FEV, indicating that PM2.5 exposure is a risk factor for poor lung function.








