---
title: "Lab 9: SQL"
author: "Nate Truettner"
format: html
---

```{r}
#install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
doctor <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/doctor.csv")
visit <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/visit.csv")
patient <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/patient.csv")
payment <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/payment.csv")

# Copy data.frames to database
dbWriteTable(con, "doctor", doctor)
dbWriteTable(con, "visit", visit)
dbWriteTable(con, "patient", patient)
dbWriteTable(con, "payment", payment)

dbListTables(con)

```
### Exercise 1
```{sql connection=con}
SELECT doctor_id, first_name, last_name
FROM doctor
ORDER by last_name, first_name
```
### Exercise 2
```{sql connection=con}
SELECT doctor_id, first_name, last_name
FROM doctor 
WHERE last_name IN ('WILLIAMS', 'DAVIS')
```
### Exercise 3
```{sql connection=con}
SELECT DISTINCT patient_id
FROM visit
WHERE date(visit_start_date) = '2005-07-05'
```

### Exercise 4
#### Exercise 4.1
```{sql connection=con}
SELECT *
FROM payment
WHERE amount IN (199, 799, 999)
```
#### Exercise 4.2
```{sql connection=con}
SELECT *
FROM payment
WHERE amount > 500
```
#### Exercise 4.3
```{sql connection=con}
SELECT *
FROM payment
WHERE amount > 500 AND amount < 800
```
### Exercise 5
```{sql connection=con}
SELECT payment.payment_id, payment.amount
FROM payment
  INNER JOIN patient ON patient.patient_id = payment.patient_id
WHERE patient.last_name == "DAVIS"
```
### Exercise 6
#### Exercise 6.1
```{sql connection=con}
SELECT COUNT(*) AS "Total Number of Visits"
FROM visit
```
#### Exercise 6.2
```{sql connection=con}
SELECT patient_id AS "Patient ID", COUNT(visit_id) AS "Number of Visits"
FROM visit
GROUP BY patient_id
```
#### Exercise 6.3
```{sql connection=con}
SELECT patient_id AS "Patient ID", COUNT(visit_id) AS "Number of Visits"
FROM visit
GROUP BY patient_id
ORDER BY COUNT(visit_id) DESC
```
#### Exercise 6.4
```{sql connection=con}
SELECT patient_id AS "Patient ID", COUNT(visit_id) AS "Number of Visits"
FROM visit
GROUP BY patient_id
HAVING count(visit_id) >= 40
ORDER BY COUNT(visit_id) DESC
```
### Exercise 7
```{sql connection=con}
SELECT MAX(amount) AS "Maximum Payment", MIN(amount) AS "Minimum Payment", AVG(amount) AS "Average Payment", SUM(amount) AS "Total Sum of Payments"
FROM payment
```

#### Exercise 7.1
```{sql connection=con}
SELECT patient_id AS "Patient ID", MAX(amount) AS "Maximum Payment", MIN(amount) AS "Minimum Payment", AVG(amount) AS "Average Payment", SUM(amount) AS "Total Sum of Payments"
FROM payment
GROUP BY patient_id
```
#### Exercise 7.2
```{sql connection=con}
SELECT patient_id AS "Patient ID", MAX(amount) AS "Maximum Payment", MIN(amount) AS "Minimum Payment", AVG(amount) AS "Average Payment", SUM(amount) AS "Total Sum of Payments"
FROM payment
GROUP BY patient_id
HAVING COUNT(payment_id) > 5
```
### Exercise 8
#### Exercise 8.1
```{sql connection=con}
SELECT doctor.doctor_id, doctor.first_name, doctor.last_name, AVG(julianday(visit.visit_end_date) - julianday(visit.visit_start_date)) AS "Average Visit Length"
FROM doctor
INNER JOIN visit
    ON doctor.doctor_id = visit.doctor_id
GROUP BY doctor.doctor_id, doctor.first_name, doctor.last_name
```
#### Exercise 8.2
```{sql connection=con}
SELECT doctor.doctor_id, doctor.first_name, doctor.last_name, AVG(julianday(visit.visit_end_date) - julianday(visit.visit_start_date)) AS "Average Visit Length"
FROM doctor
INNER JOIN visit
    ON doctor.doctor_id = visit.doctor_id
GROUP BY doctor.doctor_id, doctor.first_name, doctor.last_name
ORDER BY "Average Visit Length" DESC
LIMIT 5
```
The doctors with the 5 longest average visit lengths are Ben Harris, Tom Cruise, Natalie Hopkins, Elvis Marx, and Liza Bergman.

```{sql connection=con}
SELECT doctor.doctor_id, doctor.first_name, doctor.last_name, AVG(julianday(visit.visit_end_date) - julianday(visit.visit_start_date)) AS "Average Visit Length"
FROM doctor
INNER JOIN visit
    ON doctor.doctor_id = visit.doctor_id
GROUP BY doctor.doctor_id, doctor.first_name, doctor.last_name
ORDER BY "Average Visit Length" ASC
LIMIT 5
```

The doctors with the 5 shortest average visit lengths are Ellen Presley,  Debbie  Akroyd, Christian Gable, Vivien Bergen, and Renee Ball.

### Cleanup
```{r}
dbDisconnect(con)
```

